---
title: "lab07_handin"
format:
  html:
    embed-resources: true
editor: visual
---

## **Info**

**Group number:** 06

**Group members:**

Alberte Englund, s215067

Mathilde Due, s215063,

Line Winther Gormsen, s215111

Sigrid Frandsen, s205875

Kristine Johansen, s215098

# 1. Aim

# 2. **Load packages**

```{r}
library("tidyverse")
```

To carry out the analysis, we used packages from the `tidyverse` collection along with `broom` for tidying model outputs and `cowplot` for plot themes.

```{r}
#| message: false
#| warning: false
library("tidyverse")
library("broom")  # devtools::install_github("tidymodels/broom")
library("cowplot")
```

# 3. **Load data**

We use the `biopsy` dataset, available online through the ... repository. **The code below downloads the data and store it locally the first time, the script is run. In subsequent runs the data is only reloaded.**

```{r}
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")

biopsy
```

# 4. **Data preparation**

```{r}
pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  prcomp(scale = TRUE) # do PCA on scaled data
```

```{r}
pca_fit <- biopsy %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>% # scale data
  prcomp() # do PCA
```

# 5. **Analysis**

```{r}

pca_fit %>%
  augment(biopsy) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#D55E00", benign = "#0072B2")
  ) +
  theme_half_open(12) + background_grid()
pca_fit
```

```{r}
# extract rotation matrix
pca_fit %>%
  tidy(matrix = "rotation")
```

```{r}

# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit %>%
  tidy(matrix = "rotation") %>%
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
pca_fit
```
