---
title: "lab07_handin"
format:
  html:
    embed-resources: true
editor: visual
---

## **Info**

**Group number:** 06

**Group members:**

Alberte Englund, s215067

Mathilde Due, s215063,

Line Winther Gormsen, s215111

Sigrid Frandsen, s205875

Kristine Johansen, s215098

# 1. Aim

# 2. **Load packages**

```{r}
library("tidyverse")
```

To carry out the analysis, we used packages from the `tidyverse` collection along with `broom` for tidying model outputs and `cowplot` for plot themes.
>>>>>>> 83cc951476878a7f1e2c9a3d27674b85a38b5be5

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(broom)  # devtools::install_github("tidymodels/broom")
library(cowplot)
```

# 3. **Load data**

We use the `biopsy` dataset, available online through the ... repository. **The code below downloads the data and store it locally the first time, the script is run. In subsequent runs the data is only reloaded.**

```{r}
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")
```

# 4. **Data preparation**

# 5. **Analysis**
