---
title: "R note collection - 2025"
format: html
---

# Loading packages

## Tidyverse

*Description of tidyverse...*

```{r}
library("tidyverse")
```

Packages that are a part of tidyverse:

-   `ggplot2`
-   `dplyr`
-   `tidyr`
-   `readr`
-   `purrr`
-   `tibble`
-   `stringr`
-   `forcats`
-   `lubridate`

# Loading data

Mangler lige at loade data ind, så vi kan køre chunksne

# Data wrangling

## Join

### Full join

### Inner join

### Right/left join

### Anti join

Joins everything in the first dataset, that is not overlapping with the second dataset

```{r}
anti_join(X, Y, by = "id")
```

## Table 1 - cohort presentation

The table 1 command creates the classic table known from scientific papers as one of the first tables, which presents the cohort and its characteristics like gender, age etc.

```{r}
library("table1") # <= Yes, this should normally go at the beginning!
meta_data |>
  mutate(Gender = factor(Gender),
         Cohort = factor(Cohort)) |>
  table1(x = formula(~ Gender + Age + Race | Cohort),
         data = _)
```

# Data visualisation

# Functional programming

To make functional programming you are using the purrr package (which is a part of tidyverse). The advantage of functional programming is that you can define a function one time and if you have to make changes it only needs to be changed once.

## Double a value

Creating the function:

```{r}
double_the_value <- function(value){
  return( 2*value )
}

x <- 2
double_the_value(value = x)
```

Using the function:

```{r}
y <- double_the_value(value = x)
y
```

# Different general commands

### Separator

When you want to actually separate where the '+' is placed and not on the meaning of the '+', you need to specify it with an escape '\\\\'.

```{r}
sep = "\\+" 
```

## Slice sample

Takes out a specific number of rows from the dataset but randomly. If you only view the top or the bottom there is a risk that the data are sorted and you will only get a wrong understanding of the trend in the data. By using slice_sample you get random rows and you might get a better understanding of the data.

```{r}
meta_data |>
  slice_sample(n = 10)
```
