---
title: "R note collection - 2025"
format: html
---

# Loading data

# Data wrangling

## Join

### Anti join

Joins everything in the first dataset, that is not overlapping with the second dataset

```{r}
anti_join(X, Y, by = "id")
```

## Table 1 - cohort presentation

The table 1 command creates the classic table known from scientific papers as one of the first tables, which presents the cohort and its characteristics like gender, age etc.

```{r}
library("table1") # <= Yes, this should normally go at the beginning!
meta_data |>
  mutate(Gender = factor(Gender),
         Cohort = factor(Cohort)) |>
  table1(x = formula(~ Gender + Age + Race | Cohort),
         data = _)
```

# Functional programming

## Call functions on objects

```{r}
double_the_value <- function(value){
  return( 2*value )
}

x <- 2
double_the_value(value = x)
```

The result can be "catched" by assigning it to a variable

```{r}
y <- double_the_value(value = x)
y
```

# Different commands

### Separator

When you want to actually separate where the '+' is placed and not on the meaning of the '+', you need to specify it with an escape '\\\\'.

```{r}
sep = "\\+" 
```

## Slice sample

Takes out a specific number of rows from the dataset but randomly. If you only view the top or the bottom there is a risk that the data are sorted and you will only get a wrong understanding of the trend in the data. By using slice_sample you get random rows and you might get a better understanding of the data.

```{r}
meta_data |>
  slice_sample(n = 10)
```
